\documentclass[a4paper,11pt]{article}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{caption}
\usepackage{cite}     % For numeric citations
\usepackage{url}      % Needed for proper URL handling
\usepackage{pgfplots} % For creating plots
\usepackage{tikz}     % Required for pgfplots
\usepackage{hyperref}  % Load hyperref last (it includes url functionality)

\pgfplotsset{compat=1.16} % Set compatibility mode for pgfplots

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    citecolor=blue,        % Set citation numbers to blue color
    pdftitle={PID Control and Auto-Tuning in the VentCon Pressure Control System},
    pdfauthor={VENTREX},
}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.95}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle}

\title{PID Control and Auto-Tuning in the VentCon Pressure Control System}
\author{VENTREX}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This document describes the implementation of the PID (Proportional-Integral-Derivative) control system and auto-tuning procedures in the VentCon pressure control system. The document details the core control algorithm, anti-windup mechanisms, hysteresis compensation, and the relay-based auto-tuning method with multiple tuning rules for different performance requirements.
\end{abstract}

\tableofcontents

\section{Introduction}

The VentCon2 system is designed to provide precise pressure control using a solenoid valve operated via PWM (Pulse Width Modulation). The system employs a digital PID controller to maintain the desired pressure setpoint by dynamically adjusting the valve opening based on real-time pressure measurements. To simplify the tuning process, an automated self-tuning procedure based on the relay method is implemented.

\section{PID Control Implementation}

\subsection{Basic PID Theory}

The PID control algorithm calculates the control output $u(t)$ based on the error $e(t)$ between the desired setpoint and the measured process variable\cite{wiki:pid}:

\begin{equation}
u(t) = K_p e(t) + K_i \int_{0}^{t} e(\tau) d\tau + K_d \frac{de(t)}{dt}
\end{equation}

Where:
\begin{itemize}
    \item $K_p$ is the proportional gain
    \item $K_i$ is the integral gain
    \item $K_d$ is the derivative gain
    \item $e(t) = \text{setpoint} - \text{measured value}$
\end{itemize}

\subsection{Implementation in VentCon2}

The VentCon2 system uses the PID\_v2 library for Arduino which implements a discrete form of the PID equation. The controller is configured with the following parameters:

\begin{lstlisting}[language=C++, caption=PID Configuration]
// PID initialization
PID pid(&pressureInput, &pwmOutput, &settings.setpoint, 
        settings.Kp, settings.Ki, settings.Kd, DIRECT);
pid.SetMode(PID::Automatic);
pid.SetOutputLimits(0, pwm_max_value);
pid.SetSampleTime(SAMPLE_TIME); // 20ms (50Hz)
\end{lstlisting}

The controller updates at a frequency of 50Hz, providing a good balance between responsiveness and processing overhead. The PID output is then mapped to the valve's effective operating range (50\% to 90\% duty cycle), as the valve exhibits non-linear behavior at lower duty cycles.

\subsection{Anti-Windup Implementation}

The system implements an advanced anti-windup mechanism that prevents integral windup in two key situations:

\begin{enumerate}
    \item When the valve is below its minimum effective operating range (50\% duty cycle)
    \item When the valve is above its maximum effective operating range (90\% duty cycle)
\end{enumerate}

The anti-windup mechanism works by temporarily switching the PID controller to manual mode and back to automatic when these conditions are detected, effectively resetting the integral component:

\begin{lstlisting}[language=C++, caption=Anti-Windup Implementation]
if (settings.antiWindup) 
{
  float pidPercent = (pwmOutput / pwm_max_value) * 100.0;
  
  if ((pidPercent < VALVE_MIN_DUTY && pwmOutput > previousOutput) ||
      (pidPercent > VALVE_MAX_DUTY && pwmOutput > previousOutput)) 
  {
    // Reset the PID to prevent integral accumulation
    pid.SetMode(PID::Manual);
    pid.SetMode(PID::Automatic);
  }
}
\end{lstlisting}

\subsection{Hysteresis Compensation}

Solenoid valves typically exhibit hysteresis behavior, where the opening and closing characteristics differ. To compensate for this, the system implements a direction-based hysteresis compensation mechanism:

\begin{lstlisting}[language=C++, caption=Hysteresis Compensation]
if (settings.hysteresis) 
{
  bool currentlyIncreasing = pressureInput > lastPressure;
  
  if (currentlyIncreasing != pressureIncreasing) 
  {
    if (!currentlyIncreasing) 
    {
      // When changing direction from increasing to decreasing
      pidPercent += settings.hystAmount;
    }
    else 
    {
      // When changing direction from decreasing to increasing
      pidPercent -= settings.hystAmount;
    }
    
    pressureIncreasing = currentlyIncreasing;
  }
}
\end{lstlisting}

This approach helps overcome the "stiction" effect in solenoid valves, providing more consistent control regardless of whether the pressure is increasing or decreasing.

\section{Auto-Tuning Procedure}

\subsection{Relay Method Theory}

The auto-tuning implementation uses the relay method developed by Åström and Hägglund\cite{astrom1984automatic}, which is a variation of the Ziegler-Nichols frequency response method\cite{ziegler1942optimum}. This approach is widely used in industrial applications due to its reliability and ease of implementation\cite{wiki:pidtuning}. Instead of increasing the proportional gain until sustained oscillation occurs (which can be difficult and potentially unsafe), the relay method:

\begin{enumerate}
    \item Forces the system into controlled oscillation using a relay (on/off) controller
    \item Measures the resulting oscillation amplitude and period
    \item Calculates the "ultimate gain" ($K_u$) and "ultimate period" ($T_u$)
    \item Applies tuning rules to determine the optimal PID parameters
\end{enumerate}

The ultimate gain is calculated as:

\begin{equation}
K_u = \frac{4d}{\pi a}
\end{equation}

Where:
\begin{itemize}
    \item $d$ is the relay amplitude (AUTOTUNE\_RELAY\_STEP = 75\%)
    \item $a$ is the resulting oscillation amplitude in the process variable
\end{itemize}

\subsection{Implementation in VentCon2}

The auto-tuning process in VentCon2 works as follows:

\begin{enumerate}
    \item The system sets the target pressure to a predefined setpoint (3.0 bar)
    \item A relay controller alternates the valve between:
    \begin{itemize}
        \item 0\% (fully closed) when pressure exceeds the setpoint
        \item 75\% of effective valve opening when pressure falls below setpoint - noise band
    \end{itemize}
    \item The system records the cycle times and amplitudes for 10 complete cycles (or until timeout)
    \item The ultimate gain and period are calculated from the averaged measurements
    \item PID parameters are calculated using the selected tuning rule
\end{enumerate}

An important implementation detail is the handling of the valve's non-linear behavior. The relay amplitude (75\%) refers to the actual valve opening percentage rather than the raw PWM value. When applying the relay output in the high state, the system accounts for the valve's characteristics by mapping this percentage to the appropriate PWM value:

\begin{lstlisting}[language=C++, caption=Auto-Tuning Relay Output]
// When in high state:
float mappedDuty = VALVE_MIN_DUTY + (AUTOTUNE_RELAY_STEP_RAW / 100.0) * (VALVE_MAX_DUTY - VALVE_MIN_DUTY);
uint32_t pwmValue = (uint32_t)((mappedDuty / 100.0) * pwm_max_value);
ledcWrite(PWM_CHANNEL_MOSFET, pwmValue);

// When in low state:
ledcWrite(PWM_CHANNEL_MOSFET, 0);  // Completely closed
\end{lstlisting}

This ensures the actual valve behavior matches the theoretical model used for parameter calculation.

\subsection{Multiple Tuning Rules}

The VentCon2 system implements four different tuning rules to accommodate different control requirements\cite{wiki:pidtuning}:

\begin{table}[ht]
\centering
\caption{PID Tuning Rules\cite{ni:pidtuning}}
\begin{tabular}{lcccp{3.6cm}}
\toprule
\textbf{Tuning Rule} & \textbf{$K_p$} & \textbf{$K_i$} & \textbf{$K_d$}  \\
\midrule
\textbf{Z-N Classic} \cite{ziegler1942optimum} & $0.6K_u$ & $1.2K_u/T_u$ & $0.075K_uT_u$   \\
\multicolumn{4}{l}{Balanced response with moderate overshoot} \\
\textbf{Z-N Aggressive} & $0.7K_u \cdot \text{aggr}$ & $1.75K_u/T_u$ & $0.085K_uT_u$  \\
\multicolumn{4}{l}{Faster response with potential increased overshoot} \\
\textbf{Tyreus-Luyben} \cite{tyreus1992tuning} & $0.45K_u$ & $0.36K_u/T_u$ & $0.14K_uT_u$  \\
\multicolumn{4}{l}{Reduced overshoot, slower response} \\
\textbf{Pessen Integral} \cite{pessen1954new} & $0.7K_u$ & $1.75K_u/T_u$ & $0.105K_uT_u$  \\
\multicolumn{4}{l}{Fast setpoint tracking} \\
\bottomrule
\end{tabular}
\end{table}

Where Z-N stands for Ziegler-Nichols, and "aggr" is the aggressiveness factor (default 1.2) that can be adjusted from 0.5 to 2.0.

The system defaults to the Ziegler-Nichols Aggressive rule with an aggressiveness factor of 1.2, as this provides a good balance between response speed and stability for most applications.

\subsection{Adjustable Aggressiveness}

A unique feature of the VentCon2 auto-tuning system is the adjustable aggressiveness factor, which provides an additional layer of control over the tuning outcome:

\begin{itemize}
    \item Values from 0.5 to 1.0 provide more conservative control with less overshoot
    \item The default value of 1.2 provides faster response than classic Z-N
    \item Values from 1.5 to 2.0 generate very aggressive control for applications requiring minimal response time
\end{itemize}

This factor primarily affects the proportional gain but influences the overall system behavior, allowing fine-tuning without changing the fundamental tuning rule.

\begin{lstlisting}[language=C++, caption=Aggressiveness Implementation]
// Example of how aggressiveness affects parameter calculation
newKp = 0.7 * Ku * tuningAggressiveness;  // Proportional gain scaled by aggressiveness
\end{lstlisting}

\subsection{Auto-Tuning Workflow}

The typical auto-tuning workflow for a user is:

\begin{enumerate}
    \item Select the desired tuning rule (optional): \texttt{TUNE RULE n}
    \begin{itemize}
        \item 0: Ziegler-Nichols Classic (balanced)
        \item 1: Ziegler-Nichols Aggressive (faster response) - Default
        \item 2: Tyreus-Luyben (minimal overshoot)
        \item 3: Pessen Integral (fast setpoint tracking)
    \end{itemize}
    \item Adjust aggressiveness if needed (optional): \texttt{TUNE AGGR x}
    \item Start the auto-tuning process: \texttt{TUNE START}
    \item Wait for the process to complete (typically 1-3 minutes)
    \item Review the calculated parameters
    \item Accept or reject the parameters: \texttt{TUNE ACCEPT} or \texttt{TUNE REJECT}
\end{enumerate}

Users can view available tuning rules and the current settings with the command \texttt{TUNE RULES}.

\section{Performance Considerations}

\subsection{Tuning Rule Selection}

\begin{itemize}
    \item \textbf{Ziegler-Nichols Classic}: Provides a balanced response, but may exhibit significant overshoot
    \item \textbf{Ziegler-Nichols Aggressive}: Faster response time at the cost of potentially increased overshoot
    \item \textbf{Tyreus-Luyben}: More conservative, with reduced overshoot but slower response
    \item \textbf{Pessen Integral}: Emphasizes setpoint tracking with fast recovery from disturbances
\end{itemize}

\subsection{Comparative Step Responses}

Figure \ref{fig:step_responses} illustrates the typical step responses achieved with each tuning rule. These theoretical response curves demonstrate the relative differences in performance characteristics.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=12cm,
    height=8cm,
    xlabel={Time},
    ylabel={Pressure},
    xmin=0, xmax=10,
    ymin=0, ymax=1.5,
    xtick={0,2,4,6,8,10},
    ytick={0,0.25,0.5,0.75,1.0,1.25,1.5},
    legend pos=south east,
    grid=both,
    minor grid style={gray!25},
    major grid style={gray!25},
    title={Comparative Step Responses},
    ]
    
    % Setpoint
    \addplot[black, dashed, line width=1pt] coordinates {(0,0) (0,1) (10,1)};
    
    % Ziegler-Nichols Classic - Moderate overshoot, moderate settling
    \addplot[blue, line width=1.5pt] coordinates {
        (0,0) (0.5,0.35) (1,0.7) (1.5,0.95) (2,1.25) (2.5,1.32) (3,1.25) 
        (3.5,1.15) (4,1.05) (4.5,0.97) (5,0.98) (5.5,1.02) (6,1.03) 
        (6.5,1.01) (7,0.99) (7.5,1.0) (8,1.0) (9,1.0) (10,1.0)
    };
    
    % Ziegler-Nichols Aggressive - Higher overshoot, faster rise
    \addplot[red, line width=1.5pt] coordinates {
        (0,0) (0.25,0.22) (0.5,0.48) (0.75,0.72) (1,0.9) (1.25,1.1) 
        (1.5,1.32) (1.75,1.42) (2,1.38) (2.5,1.22) (3,1.08) (3.5,0.95) 
        (4,0.97) (4.5,1.04) (5,1.05) (5.5,1.02) (6,0.99) (6.5,0.98) 
        (7,1.0) (8,1.0) (9,1.0) (10,1.0)
    };
    
    % Tyreus-Luyben - Less overshoot, slower rise and settling
    \addplot[green!70!black, line width=1.5pt] coordinates {
        (0,0) (0.5,0.18) (1,0.35) (1.5,0.52) (2,0.68) (2.5,0.82) (3,0.92) 
        (3.5,0.97) (4,1.02) (4.5,1.05) (5,1.04) (5.5,1.02) (6,1.01) 
        (6.5,1.0) (7,1.0) (8,1.0) (9,1.0) (10,1.0)
    };
    
    % Pessen Integral - Fast rise, moderate overshoot, quick settling
    \addplot[orange, line width=1.5pt] coordinates {
        (0,0) (0.25,0.25) (0.5,0.52) (0.75,0.75) (1,0.95) (1.25,1.15) 
        (1.5,1.28) (1.75,1.35) (2,1.28) (2.25,1.15) (2.5,1.05) (2.75,0.98) 
        (3,0.97) (3.5,0.99) (4,1.01) (4.5,1.0) (5,1.0) (6,1.0) (7,1.0) 
        (8,1.0) (9,1.0) (10,1.0)
    };

    \legend{Setpoint, Z-N Classic, Z-N Aggressive, Tyreus-Luyben, Pessen Integral}
\end{axis}
\end{tikzpicture}
\caption{Typical step responses for different tuning rules applied to the pressure control system. The plot shows how each rule affects rise time, overshoot, and settling characteristics.}
\label{fig:step_responses}
\end{figure}

The plot illustrates several important characteristics:

\begin{itemize}
    \item \textbf{Z-N Classic (blue)}: Shows a moderate rise time with significant overshoot (approximately 32\%) and some oscillation before settling.
    
    \item \textbf{Z-N Aggressive (red)}: Demonstrates the fastest rise time with the highest overshoot (approximately 42\%) but can take longer to fully settle due to more pronounced oscillations.
    
    \item \textbf{Tyreus-Luyben (green)}: Exhibits the slowest rise time but with minimal overshoot (approximately 5\%), making it ideal for applications where overshoot must be minimized.
    
    \item \textbf{Pessen Integral (orange)}: Provides a fast rise time with moderate overshoot (approximately 35\%) but settles more quickly than the Ziegler-Nichols methods, making it well-suited for setpoint tracking applications.
\end{itemize}

The aggressiveness factor in the Z-N Aggressive method can shift its response curve between the Classic and Aggressive profiles shown here, providing flexible tuning options.

\subsection{Practical Recommendations}

\begin{enumerate}
    \item Start with the default (Ziegler-Nichols Aggressive, aggressiveness=1.2)
    \item If the response is too sluggish:
    \begin{itemize}
        \item First try increasing aggressiveness: \texttt{TUNE AGGR 1.5} or \texttt{TUNE AGGR 2.0}
        \item If still insufficient, try Pessen Integral rule: \texttt{TUNE RULE 3}
    \end{itemize}
    \item If overshoot is problematic:
    \begin{itemize}
        \item First try reducing aggressiveness: \texttt{TUNE AGGR 0.8}
        \item If still insufficient, switch to Tyreus-Luyben: \texttt{TUNE RULE 2}
    \end{itemize}
    \item Enable anti-windup (\texttt{AW ON}) for all tuning methods to improve recovery from saturation
    \item For valves with significant stiction, enable hysteresis compensation (\texttt{HYST ON})
    \item After auto-tuning, fine-tune individual parameters if necessary:
    \begin{itemize}
        \item To reduce overshoot: reduce Kp or increase Kd
        \item To eliminate steady-state error: increase Ki
        \item To reduce oscillation: increase Kd or reduce Kp
    \end{itemize}
\end{enumerate}

\subsection{Limitations}

The auto-tuning method has several limitations to be aware of:

\begin{itemize}
    \item Works best for linear or approximately linear systems
    \item May not produce optimal results for highly non-linear systems
    \item Performance can be affected by noise (mitigated with AUTOTUNE\_NOISE\_BAND)
    \item Requires a stable starting point for best results
    \item External disturbances during tuning can affect accuracy
\end{itemize}

\section{Conclusion}

The VentCon2 system implements a sophisticated PID control system with advanced features like anti-windup and hysteresis compensation. The automated tuning procedure based on the relay method provides an accessible way to optimize the control parameters for specific applications without requiring in-depth knowledge of control theory\cite{astrom1995pid}.

The multiple tuning rules offer flexibility to balance between response speed, overshoot, and disturbance rejection based on the specific requirements of the application\cite{odwyer2009handbook}\cite{controlguru}\cite{controlnotes}\cite{oreilly:pidtuning}.

\nocite{*}  % Include all references, even if not explicitly cited

\section{References}

\begin{thebibliography}{99}

\bibitem{astrom1984automatic}
Åström, K.J. and Hägglund, T. (1984).
\newblock Automatic tuning of simple regulators with specifications on phase and amplitude margins.
\newblock \emph{Automatica}, 20(5):645--651.

\bibitem{ziegler1942optimum}
Ziegler, J.G. and Nichols, N.B. (1942).
\newblock Optimum settings for automatic controllers.
\newblock \emph{Transactions of the ASME}, 64:759--768.

\bibitem{tyreus1992tuning}
Tyreus, B.D. and Luyben, W.L. (1992).
\newblock Tuning PI controllers for integrator/dead time processes.
\newblock \emph{Industrial \& Engineering Chemistry Research}, 31(11):2625--2628.

\bibitem{pessen1954new}
Pessen, D.W. (1954).
\newblock A new look at PID-controller tuning.
\newblock \emph{Journal of Basic Engineering}, 76:494--501.

\bibitem{astrom1995pid}
Åström, K.J. and Hägglund, T. (1995).
\newblock \emph{PID Controllers: Theory, Design, and Tuning}.
\newblock Instrument Society of America, Research Triangle Park, NC, 2nd edition.

\bibitem{odwyer2009handbook}
O'Dwyer, A. (2009).
\newblock \emph{Handbook of PI and PID controller tuning rules}.
\newblock Imperial College Press, London, 3rd edition.

\bibitem{wiki:pid}
Wikipedia (2023).
\newblock PID controller.
\newblock \url{https://en.wikipedia.org/wiki/PID_controller}
\newblock [Online; accessed November 2023]

\bibitem{wiki:pidtuning}
Wikipedia (2023).
\newblock PID controller tuning.
\newblock \url{https://en.wikipedia.org/wiki/PID_controller#Loop_tuning}
\newblock [Online; accessed November 2023]

\bibitem{controlguru}
Control Guru (2023).
\newblock PID control and auto-tuning resource site.
\newblock \url{https://controlguru.com/}
\newblock [Online; accessed November 2023]

\bibitem{controlnotes}
MathWorks (2023).
\newblock PID Tuning Algorithms.
\newblock \url{https://www.mathworks.com/help/control/ug/pid-tuning-algorithms.html}
\newblock [Online; accessed November 2023]

\bibitem{oreilly:pidtuning}
Control Engineering Practice (2023).
\newblock PID Control Tuning: Principles and Applications.
\newblock \url{https://blog.opticontrols.com/archives/344}
\newblock [Online resource; accessed November 2023]

\bibitem{ni:pidtuning}
National Instruments (2023).
\newblock PID theory explained.
\newblock \url{https://www.ni.com/en-us/innovations/white-papers/06/pid-theory-explained.html}
\newblock [Online; accessed November 2023]

\end{thebibliography}

\end{document}
